---
title: "Abscessus_analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Abscessus_analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Background
 
This is a study comparing the mortality of transplant patients: those who were infected with Mycobacterium abscessus vs those who were not infected. To me, this suggests a retrospective cohort study: Patients were either exposed or unexposed (to the infection), then either did or did not experience a binary outcome (death). The resulting 2x2 table includes a cell containing only 2 patients, so I used Fisher's exact test. I hadn't realised that I would be given survival time as part of the data. I became concerned that I was throwing information away, so I then did some standard survival analysis, including a log rank test due to uncertainty over the PH assumption. This seemed sensible to me.

The paper has been reviewed and the decision is that the statisticial analysis is _flawed_! There is immortal time bias, because the exposure (infection) covariate is time-varying and not fixed. Any revision will be _rejected_ unless a landmark analysis is implemented. I accept this criticism, though I also would defend my original analysis using Fisher's exact test, as I think this is not biased. 

I have never done a landmark analysis (thanks for your survival analysis book, Dom!). My understanding is that to do a landmark analysis, I need to choose a cut-off point. At this cut-off point, (1) all data before the cut-off is discarded, and (2) each patients' infection status is not allowed to change. That is, any patient that is uninfected at the cut-off point must be treated as uninfected in the analysis, regardless of what happens to them.

There are only 9 (nine) patients in the infected group. With this in mind, I think the most sensible cut-off point would be the day X after transplant, where X equals the smallest day number that would put all 9 infected patients in the infected group. I would like to hear an opinion on this. I cannot do this analysis at the moment as I don't have this data; I only have death and post-transplant survival/censoring time. I think I only need "number of days from transplant to infection" for the 9 patients to learn X. I would then discard all patients who died before day X, and use a new survival variable, "survival/censoring time from day X".

I don't want the paper to be rejected for bad stats(!), so I would really appreciate an opinion on this before I go ahead.


## Analysis (as submitted)

### Load packages, load anonymised data, tidy data

```{r setup, echo=T}
library(librarian)
librarian::shelf(survival, survminer,martinlaw/abscessus)

dat <- read.csv("../data/abscessus_data.csv")

# Create more informative variables:
dat$infected <- ifelse(test=dat$abscessuss_yes_1, yes="Infected", no="Not infected")
dat$alive <- ifelse(test=dat$alive_yes_1, yes="Alive", no="Not alive")
dat$lung.dysfun <- ifelse(test=dat$CLAD_yes_1, yes="Yes", no="No")
dat$airway.injury <- ifelse(test=dat$airway_injury_yes_1, yes="Injury", no="No Injury")

names(dat)[names(dat)=="survival_post_tx"] <- "time"
dat$status <- -1*(dat$alive_yes_1-1)
```

## Difference in mortality
### Binary
Begin with 2x2 table:
```{r, binary mortality}
con <- table(dat$infected, dat$alive)
con
```

Treating survival as binary: No difference between groups when using Fisher's Exact Test:

```{r}
fisher.test(con) 
```

### Survival analysis
#### Plot
```{r plot mortality, fig.height=6, fig.width=8}
fit <- survfit(Surv(time, status) ~ infected, data = dat)
#ggsurvplot(fit, data=dat)
saved_theme <- theme_minimal() + theme(axis.text = element_text(size = 10))
logrank.plot <- ggsurvplot(
   fit,                     # survfit object with calculated statistics.
   data = dat,              # data used to fit survival curves. 
   risk.table = TRUE,       # show risk table.
   pval = TRUE,             # show p-value of log-rank test.
   conf.int = FALSE,         # show confidence intervals for 
                            # point estimates of survival curves.
   #xlim = c(0,2000),        # present narrower X axis, but not affect
                            # survival estimates.
   #break.time.by = 500,     # break X axis in time intervals by 500.
   ggtheme = saved_theme, # customize plot and risk table with a theme.
 risk.table.y.text.col = T, # colour risk table text annotations.
  risk.table.y.text = FALSE, # show bars instead of names in text annotations
                            # in legend of risk table
 title="Kaplan-Meier survival curves",
)
logrank.plot
png(filename="../figs/logrank_plot.png", width=8, height=6, units="in", res=300)
logrank.plot
dev.off()
```

#### Log-rank test
The log-rank test finds no difference in survival times between the two groups, and is the most appropriate model as it does not assume proportional hazards.
```{r logrank}
fit.diff <- survdiff(formula(fit), data=dat)
fit.diff
# P value:
fit.diff$pvalue

obs.exp.ratios <- fit.diff$obs/fit.diff$exp
# Hazard ratio:
obs.exp.ratios[1]/obs.exp.ratios[2]

# Confidence interval
obs.minus.exp <- fit.diff$obs[1]-fit.diff$exp[1]
V <- fit.diff$var[1,1]
# V <- obs.minus.exp^2 / fit.diff$chisq # equivalent calculation
K <- obs.minus.exp/V
log.lower <- K-1.96*sqrt(V)
log.upper <- K+1.96*sqrt(V)
ci <- exp(c(log.lower, log.upper))
ci
```
#### Cox PH
Fitting a Cox proportional hazards model also does not show an effect, though from the plot above, the assumption of proportional hazards appears to not hold.
```{r coxPH}
cox.no.covar <- coxph(Surv(time, status) ~ infected, data = dat)
cox.no.covar
```

### Descriptive statistics
Using Wilcoxon rank-sum test for continuous variables and Fisher's Exact Test for categorical variables (due to the small group size)

#### Survival/censoring time, age at transplant
```{r descriptive stats}
# Follow up:
summary(dat$time)

wilcox.test(formula = age_at_tx ~ infected, data=dat)
tapply(dat$age_at_tx, dat$infected, summary)
```

#### BMI pre-transplant
```{r bmi}
wilcox.test(formula = bmi_pre_tx ~ infected, data=dat)
tapply(dat$bmi_pre_tx, dat$infected, summary)
```

#### Sex
```{r sex}
fisher.test(dat$infected, dat$gender_male_1)
round(prop.table(table(dat$infected, dat$gender_male_1), margin=1), 2)
```

#### Donor type
```{r donor}
fisher.test(dat$infected, dat$donortype)
round(prop.table(table(dat$infected, dat$donortype), margin=1), 2)
```

#### Organ infected
```{r organ}
fisher.test(dat$infected, dat$organ_1_single)
round(prop.table(table(dat$infected, dat$organ_1_single), margin=1), 2)
```

#### GORD
```{r GORD}
fisher.test(dat$infected, dat$gord_yes_1)
round(prop.table(table(dat$infected, dat$gord_yes_1), margin=1), 2)
```

#### Immunosuppressed
```{r immuno}
fisher.test(dat$infected, dat$immunosuppressed_pre_yes_1)
round(prop.table(table(dat$infected, dat$immunosuppressed_pre_yes_1), margin=1), 2)
```

#### PGD
```{r pgd}
fisher.test(dat$infected, dat$PGD_yes_1)
# Note infinite OR due to zero cell.
round(prop.table(table(dat$infected, dat$PGD_yes_1), margin=1), 2)
```





