---
title: "Abscessus_analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Abscessus_analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

### Load packages and anonymised data, clean 

```{r setup}
library(abscessus)
librarian::shelf(survival, survminer)

dat <- read.csv("../data/abscessus_data.csv")

# Create more informative variables:
dat$infected <- ifelse(test=dat$abscessuss_yes_1, yes="Infected", no="Not infected")
dat$alive <- ifelse(test=dat$alive_yes_1, yes="Alive", no="Not alive")
dat$lung.dysfun <- ifelse(test=dat$CLAD_yes_1, yes="Yes", no="No")
dat$airway.injury <- ifelse(test=dat$airway_injury_yes_1, yes="Injury", no="No Injury")

names(dat)[names(dat)=="survival_post_tx"] <- "time"
dat$status <- -1*(dat$alive_yes_1-1)
```

## Difference in mortality
### Binary
Begin with 2x2 table:
```{r, binary mortality}
con <- table(dat$infected, dat$alive)
con
```

Treating survival as binary: No difference between groups when using Fisher's Exact Test:

```{r}
fisher.test(con) 
```

### Survival analysis
#### Plot
```{r plot mortality, fig.height=4, fig.width=6}
fit <- survfit(Surv(time, status) ~ infected, data = dat)
ggsurvplot(fit, data=dat)
```

#### Log-rank test
The log-rank test finds no difference in survival times between the two groups, and is the most appropriate model as it does not assume proportional hazards.
```{r logrank}
fit.diff <- survdiff(formula(fit), data=dat)
fit.diff
fit.diff$pvalue
```
#### Cox PH
Fitting a Cox proportional hazards model also does not show an effect, though from the plot above, the assumption of proportional hazards appears to not hold.
```{r coxPH}
cox.no.covar <- coxph(Surv(time, status) ~ infected, data = dat)
cox.no.covar
```





